---
title: "Peer Assesment 1"
output: html_document
---

This is my work for peer assessment 1, wherein I created summary graphics for the data from an activity tracker.  The data comes from a single subject who wore an activity tracker for a 2-month period.  The tracker provided data that is broken into 5-minute intervals for each day in the dataset.

###First Task: What is mean total number of steps taken per day?

Opening the data.
```{r openfile,echo=TRUE}
activity <- read.csv("activity.csv")
```

1.) My first step is to use the dplyr package to group the database by day and then create a data frame that contains the total number of steps for each day
```{r by_day,echo=TRUE}
activity <- read.csv("activity.csv")

library(dplyr)
#Creating the data frame:
by_day <- group_by(activity, date)
Daily_Steps <- summarise(by_day, steps = sum(steps, na.rm=TRUE))
```

2.) Now to plot the daily step totals in a histogram. I am using ggplot2 to plot the data and then using the scales package to make the break points on the y-axis cleaner. As can be seen in the histogram, there are at least 10 days with either no data or minimal data.
```{r first_histo,echo=TRUE}
library(ggplot2)
library(scales)
ggplot(data=Daily_Steps, aes(steps)) +
  geom_histogram(breaks=seq(0, 25000, by =500), 
		aes(fill=..count..),
		alpha=.8) +
	labs(title="Histogram of Daily Step Count") +
	labs(x="Number of Steps", y="Number of Days") +
	ylim(c(0,10)) +
	scale_y_continuous(breaks= pretty_breaks())
```

3.) The below code finds the mean and median number of steps each day
```{r mean_median,echo=TRUE}
mean_steps <- mean(Daily_Steps$steps, na.rm=TRUE)
mean_steps <- format(round(mean_steps, 2), nsmall = 2)
median_steps <- median(Daily_Steps$steps, na.rm=TRUE)
```

Across the two months of data, the average number of steps the subject took was `r mean_steps`, while the median number of steps was `r median_steps`.



###Second Task: What is the average daily activity pattern?



1.) In order to create a time-series plot of the mean number of steps across all 5-minute intervals, I am again using the dplyr package.  This time I am grouping the database by interval and then finding the average number of steps in those intervals across 60 days in the dataframe.  I am also using ggplot to create the time-series plot.

```{r interval_timeseries,echo=TRUE}
#finding the average acorss interval
by_interval <- group_by(activity, interval)
Interval_Steps <- summarise(by_interval, steps_avg = mean(steps, na.rm=TRUE))

#Now to plot the average number of steps taken over the entire Two month period
#in each 5-minute interval period
ggplot(Interval_Steps, aes(x=interval, y=steps_avg)) +
    geom_line(aes(colour=steps_avg)) +
  scale_x_continuous(breaks=c(0,300,600,900,1200,1500,1800,2100,2400)) +
	labs(title="Average Number of Steps by 5-Minute Interval") +
	labs(x="Time(24-Hour Clock)", y="Number of Steps")
```

2.) The code below find the 5-minute interval, on average across all the days in the dataset, that contains the maximum number of steps.
```{r max_step,echo=TRUE}
max_interval <- Interval_Steps$interval[which.max(Interval_Steps$steps)]
```

The interval with the greatest average number of steps occured at `r max_interval` AM.




###Third Task: Imputing Missing Values


1.) Here is how I found the total number of missing values in the dataset
```{r Tot_missing,echo=TRUE}
tot_missing <- length(which(is.na(activity$steps)))
```

In total, there are `r tot_missing` missing values in the data set (a missing value is a 5-minute interval without an observation).


2.) & 3.) My strategy to fill-in the missing interval values was to find the average number of steps in that given interval across the 2 months. I am not using the average daily number of steps because of what can be seen in the above histogram: there are quite a few days that are entirely missing or have very few observations. In order to do this, I will start by using the "interval average" variable I created in Task #2 (I called it steps_avg). Dplyr will allow me to join the original dataframe with the dataframe containg the "interval average" in order to replace all missing intervals with that interval's 2-month average.

```{r imputation,echo=TRUE}
colnames(Interval_Steps)[2] <- "steps_avg"
activity2 <- left_join(activity, Interval_Steps, by = "interval")
activity2$steps[is.na(activity2$steps)] <- activity2$steps_avg[is.na(activity2$steps)]
```

4.) I am now re-doing the histogram from Task #1, but using the data with the imputed missing values.  I am re-doing my work from Task 1 here.

```{r imputed_histo,echo=TRUE}
#Fidning the new averages and creating the new dataframe
by_day2 <- group_by(activity2, date)
Daily_Steps2 <- summarise(by_day2, steps = sum(steps, na.rm=TRUE))


#Now to plot the daily step totals in a histogram
ggplot(data=Daily_Steps2, aes(steps)) +
  geom_histogram(breaks=seq(0, 25000, by =500), 
		aes(fill=..count..),
		alpha=.8) +
	labs(title="Histogram of Daily Step Count") +
	labs(x="Number of Steps", y="Number of Days") +
	ylim(c(0,12)) +
	scale_y_continuous(breaks= pretty_breaks())

#Finding the mean and median of steps per day, they are now larger after 
#sustituting for the NAs
imputed_mean <- mean(Daily_Steps2$steps, na.rm=TRUE)
imputed_mean <- format(round(imputed_mean, 2), nsmall = 2)
imputed_median <- median(Daily_Steps2$steps, na.rm=TRUE)
imputed_median <- format(round(imputed_median, 2), nsmall = 2)
```


After imputing missing values, the overall mean number of daily steps increased from `r mean_steps` to `r imputed_mean`, while the median number of steps increased from `r median_steps` to `r imputed_median`.  This is an expected finding given that we earlier found there to be 10 days with missing/minimal data in the original dataframe.  The new histogram shows that the number of days with missing/minimal data decresed from 10 to 2.




###Fourth Task: Graphing the difference in activity between weekdays and weekends.



1.) Below is my code to create the factor variable indicating if the day is on the weekend or weekday

```{r weekend,echo=TRUE}
#usign weekday() to indicate the day of the week
activity2$Day <- weekdays(as.Date(activity2$date))
activity2$sat <- as.numeric(activity2$Day=="Saturday")
activity2$sun <- as.numeric(activity2$Day=="Sunday")
activity2$weekend <- activity2$sat + activity2$sun
activity2$sat <- NULL
activity2$sun <- NULL

#Cretign a df for weekend and weekdays, I doing this in order to complete the next step
week_end <- subset(activity2, weekend==1)
week_day <- subset(activity2, weekend==0)
```

2.) Below I am creating the two panel plots to show the difference in activities across the weekend and weekday.  As can be seen in the plot, there is quite a difference in walking pattern between the weekend and weekdays.  The person wearing the watch tends to not rush as much in morning on the weekends and walks a little more during the day time.  On weekdays, he clearly does a bit of walking in the morning (possibly to get to work or class) and is sedentary during the day.

```{r week_plot,echo=TRUE}
#First creating the data frames of steps averaged by days during the week and
#days during the weekend, and re-merging back together 
week_day_interval <- group_by(week_day, interval)
week_day_steps <- summarise(week_day_interval, steps_avg = mean(steps, na.rm=TRUE))
week_day_steps$weekend <- "Weekday"

week_end_interval <- group_by(week_end, interval)
week_end_steps <- summarise(week_end_interval, steps_avg = mean(steps, na.rm=TRUE))
week_end_steps$weekend <- "Weekend"

week_steps <- merge(week_day_steps, week_end_steps, all=TRUE)

#Now to create the 2 separate plots
ggplot(data = week_steps, aes(interval, steps_avg)) +
     geom_line(aes(colour=steps_avg)) +
     facet_grid(weekend ~ .) +
  scale_x_continuous(breaks=c(0,300,600,900,1200,1500,1800,2100,2400)) +
	labs(title="Average Number of Steps by 5-Minute Interval") +
	labs(x="Time(24-Hour Clock)", y="Number of Steps")
```